---
title: "End of L2 exercise"
author: "Geoffrey Millard"
date: "9/13/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Retrieving USGS and EPA data

The `dataRetrieval` package works very well if you know the site numbers from which you want to pull data.  You can also get information about the data available at a site and detailed location information.  We are interested in pulling data from two sites for day two of our exercise: `c('0134277112', '0134277114')`. 

Site 112 was treated with Calcium Carbonate on October 1st 2013 and site 114 is a nearby reference site.

We don't really want all the data from these two sites, so we can specify a few different parameters `c('00403', '00409','00681', '00945', '50287', '50285')`: pH, Alkalinity, Dissolved Organic Carbon, Sulfate, Mercury, Methylmercury.

The default NWIS import will keep the parameter codes, which are not incredibly helpful, so you are going to want to replace the code with an actual name.  While you're at it, might as well replace the site codes with a more descriptive name as well.

We don't need to keep all these columns, we are really only interested in: site_no, sample_dt, sample_tm, parm_cd and result_va.  Can you make a new df with just these variables?

**Action items**

- replace site_no and parm_cd with descriptive names

- create a new, simplified dataframe



*Bonus Challenge* This dataset is a pre/post Treatment response.  Can you add in a treatment column based on the sample_dt? 

Treatment Period | Date Range 
---------------- | ----------------------
pre-treatment    | < 2013-10-1         
transitional     | 2013-10-1 < 2014-4-1
post-treatment   | > 2014-4-1


Here is some code to retrieve the data we want:

```{r pull from NWIS, warning=FALSE}
#USGS and EPA dataset retrieval package
# install.packages("dataRetrieval")
library(dataRetrieval)

#click on "Packages" tab to see functions available in the dataRetrival package

#these are USGS sites upstream from Honnedaga Lake, NY 
sites <- c('0134277112', '0134277114')

#detailed location information about the sites
# readNWISsite(c('0134277112', '0134277114'))

#whatNWISdata displays the datasets available at each site
available <- whatNWISdata(siteNumber=c('0134277112', '0134277114'))
actual <- readNWISpCode(parameterCd = available$parm_cd) #interpret parameter codes
want <- c('00403', '00409','00681', '00945', '50287', '50285') #we want: pH, Alkalinity, DOC, DSO4, DHg and DMeHg
data <- readNWISqw(siteNumbers = sites, parameterCd=want) #get the data
codes <- readNWISpCode(parameterCd = unique(data$parm_cd)) #confirms that we got what we wanted, shows UNITS! 
```


```{r, include=FALSE}
library(tidyverse)
library(lubridate)

data1 <- data

data1 <- data1 %>% mutate(Analyte = recode(parm_cd, 
                         '00403'  = 'pH',
                         '00409'  = 'Alkalinity',
                         '00681' = "DOC", 
                         '00945' = "SO4", 
                         '50287' = "Hg", 
                         '50285' = "MeHg"))

data1 <- data1 %>% mutate(Site = recode(site_no, 
                         '0134277112' = "Reference", 
                         '0134277114' = "Treated"))
colnames(data1)
data1 <- data1 %>% dplyr::select(Site, Date=sample_dt, Time=sample_tm, Analyte, Result=result_va)

data1$Treatment <- NA
data1$Treatment[ data1$Date < ymd('2013-10-1')  ] <- 1
data1$Treatment[ data1$Date >= ymd('2013-10-1') ] <- 2
data1$Treatment[ data1$Date > ymd('2014-02-28') ] <- 3  # error because I am using the wrong lubridate function
```

